Simulating High-Dimensional Multivariate Data

### 2020-12-16 010-background.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1

Moreover, the simulation method must **scale** to high dimensions:

@@

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This provides a powerful, natural way to characterize joint probability structures.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

>>>>

Moreover, the simulation method must *scale* to high dimensions:

@@

Both $\tau$ and $\rho_S$ are *invariant to under monotone transformations* of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

There is a strong connection of our simulation strategy to Gaussian *copulas* (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This provides a powerful, natural way to characterize joint probability structures.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

### 2020-12-16 020-simulation-algorithm.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$.
Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
The method is best understand as a **high-performance Gaussian copula** (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

@@

1. Simulate a random vector $\bf Z$ with $d$ **independent** and **identical** standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.

>>>>

This section describes our methods involved in simulating a random vector $\bf Y$ with $Y_i$ components for $i=1,2,\ldots,d$.
Each $Y_i$ has a specified marginal CDF $F_i$ and its inverse $F^{-1}_i$.
To characterize dependency, every pair $(Y_i, Y_j)$ has a given Pearson correlation $\rho_P$, Spearman correlation $\rho_S$, and/or Kendall's $\tau$.
The method is best understand as a *high-performance Gaussian copula* (Equation \@ref(eq:gauss)) providing a high-dimensional NORTA-inspired algorithm.

@@

1. Simulate a random vector $\bf Z$ with $d$ *independent* and *identical* standard normal components.
2. Determine the input matrix $\Sigma_{\bf Z}$ that corresponds with the specified output $\Sigma_{\bf Y}$.
3. Produce a Cholesky factor $M$ of $\Sigma_{\bf Z}$ such that $M M^{\prime}=\Sigma_{\bf Z}$.
4. Set $X$ by $X \gets MZ$.
5. $\text{Return} \; Y \; \text{where} \; Y_i \gets F_{Y_i}^{-1}[\Phi(X_i)], \; i=1,2,...,d$.

### 2020-12-16 030-bigsimr-package.Rmd
4th draft. Fix the (X, e^X) bound calculation
4th draft. Add packages.bib to address WWP ch030 part 1

The `bigsimr` package is a high-performance implementation of the proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms) for details).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality`.
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
For this low-dimensional setting, we compute using a single central processing unit (CPU) as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package.

@@

Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).

>>>>

The `bigsimr` package is a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

This subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the often-used example data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization.
For this low-dimensional setting, we operate using a single central processing unit (CPU), as the overhead in forking the tasks to multiple cores outweighs the computational gains. 
Then we transition to advanced use where we briefly describe some of the high-performance features and syntax.
The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package [@R-rslurm].

@@

Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
On the other hand, the bounds on the Pearson correlation coefficient $\rho_P$ between these margins is

### 2020-12-16 index.Rmd
4th draft. Add packages.bib to address WWP ch030 part 1

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

>>>>

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'rslurm'
), 'packages.bib')
```
